---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# JointODE

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental) [![R-CMD-check](https://github.com/ziyangg98/JointODE/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ziyangg98/JointODE/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/ziyangg98/JointODE/graph/badge.svg)](https://app.codecov.io/gh/ziyangg98/JointODE)

<!-- badges: end -->

The **JointODE** package provides a unified framework for joint modeling of longitudinal biomarker measurements and time-to-event outcomes using ordinary differential equations (ODEs).

## Overview

JointODE models biomarker evolution using **ordinary differential equations** (ODEs) and links them to survival outcomes. Unlike traditional approaches using splines or polynomials, ODEs naturally capture the dynamic behavior of biomarker trajectories.

### Model Structure

**Longitudinal Model:** Biomarker trajectories evolve according to a second-order ODE:

$$\ddot{m}_i(t) + 2 \xi_i \omega_i \dot{m}_i(t) + \omega_i^2 m_i(t) = k \omega_i^2 \mu_i(t)$$

where $m_i(t)$ is the biomarker value, $\dot{m}_i(t)$ is velocity (rate of change), and $\ddot{m}_i(t)$ is acceleration. Subject-specific dynamics are characterized by natural frequency $\omega_i$ and damping ratio $\xi_i$, with $\mu_i(t)$ representing external forcing (e.g., treatment effects, covariates). Individual heterogeneity is captured through random effects on these ODE parameters.

**Survival Model:** The hazard function incorporates biomarker dynamics:

$$\lambda_i(t) = \lambda_{0}(t)\exp\left[\alpha_1 m_i(t) + \alpha_2 \dot{m}_i^{\gamma}(t) + \mathbf{W}_i^{\top}\boldsymbol{\phi}\right]$$

where $\gamma \in \{0, 1, 2\}$ controls the velocity power (0: no velocity, 1: linear, 2: quadratic).

### Key Features

- **Subject-specific dynamics**: Random effects on ODE parameters capture individual heterogeneity
- **Flexible hazard**: B-spline baseline hazard with biomarker value and velocity effects
- **Efficient computation**: C++ implementation with parallel processing support

For full mathematical details, see the [technical documentation](http://gongziyang.com/JointODE/articles/technical-details.html).

## Installation

You can install the development version of JointODE from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("ziyangg98/JointODE")
```

## Quick Start

Here's a basic example using the included simulated dataset:

```{r example}
library(JointODE)

# Load example dataset (200 subjects with longitudinal and survival data)
data(sim)

# Fit joint ODE model
longitudinal_data <- sim$data$longitudinal_data[
  , c("id", "time", "observed", "x1", "x2")
]
t0 <- proc.time()
fit <- JointODE(
  longitudinal_formula = observed ~ biomarker + velocity + x1 + x2 +
    (biomarker + velocity | id),
  survival_formula = Surv(time, status) ~ w1 + w2,
  longitudinal_data = longitudinal_data,
  survival_data = sim$data$survival_data,
  control = list(parallel = TRUE, verbose = 3)
)
cat(sprintf("Elapsed: %.1f s\n", (proc.time() - t0)["elapsed"]))
```

```{r output}
# Model summary
summary(fit)

# Plot results
plot(fit)
```

The formula specifies:

- **ODE terms**: `biomarker` and `velocity` are the state variables (value and slope) in the ODE
- **Covariates**: `x1` and `x2` are external variables affecting the dynamics
- **Random effects**: `(biomarker + velocity | id)` allows subject-specific coefficients on the ODE value and slope terms

## Learn More

- **Getting Started**: See `vignette("JointODE")` for a detailed tutorial
- **Technical Details**: See `vignette("technical-details")` for mathematical formulations
- **Model Comparison**: See `vignette("comparison")` for comparisons with traditional joint models


## Code of Conduct

Please note that the JointODE project is released with a [Contributor Code of Conduct](http://gongziyang.com/JointODE/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
