---
title: "JointODE Examples: PBC Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{JointODE Examples: PBC Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  eval = requireNamespace("JMbayes2", quietly = TRUE),
  fig.width = 8, fig.height = 5
)
```

JointODE models biomarker trajectories with a second-order ODE and links
them to survival. This vignette walks through a practical model building
workflow using PBC serum albumin data.

# Data Preparation

```{r data-prep}
library(JointODE)
library(JMbayes2)
library(survival)
library(dplyr)
library(ggplot2)

survival_data <- pbc2.id |>
  transmute(
    id = as.integer(id), time = years, status = status2,
    drug = as.numeric(drug == "D-penicil"),
    age = scale(age)[, 1],
    sex = as.numeric(sex == "female")
  )

longitudinal_data <- pbc2 |>
  transmute(id = as.integer(id), time = year, observed = log(albumin)) |>
  left_join(survival_data |> select(id, drug, age, sex), by = "id")

# Standardize --- ODE parameters scale with the response
alb_mean <- mean(longitudinal_data$observed)
alb_sd <- sd(longitudinal_data$observed)
longitudinal_data$observed <- (longitudinal_data$observed - alb_mean) / alb_sd

cat(sprintf(
  "Subjects: %d | Events: %d (%.0f%%) | Observations: %d\n",
  nrow(survival_data), sum(survival_data$status),
  100 * mean(survival_data$status), nrow(longitudinal_data)
))
```

```{r trajectories, fig.height=6}
set.seed(42)
sample_ids <- sample(unique(longitudinal_data$id), 20)

longitudinal_data |>
  filter(id %in% sample_ids) |>
  left_join(survival_data |> select(id, status), by = "id") |>
  ggplot(aes(time, observed, group = id, color = factor(status))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1, alpha = 0.6) +
  scale_color_manual(
    values = c("0" = "steelblue", "1" = "tomato"),
    labels = c("Censored", "Event"), name = "Status"
  ) +
  labs(x = "Time (years)", y = "Log Albumin (standardized)") +
  theme_minimal(base_size = 12)
```

# Random Effects Selection

The ODE is $m''(t) = b_1 m(t) + b_2 m'(t) + f(t)$, where $b_1$
controls natural frequency, $b_2$ controls damping, and $f(t)$ is
covariate-driven forcing. Initial state RE $(m_{0,i}, v_{0,i})$ are
always included. We use `MarginalODE` to determine which additional RE
are needed.

## Three Levels of Complexity

```{r marginal-fits}
# Level 0: Fixed dynamics, random initial state only
mfit0 <- MarginalODE(
  observed ~ biomarker + velocity + drug,
  longitudinal_data,
  control = list(parallel = TRUE, verbose = 1)
)

# Level 1: + random forcing intercept (individual equilibrium levels)
mfit1 <- MarginalODE(
  observed ~ biomarker + velocity + drug + (1 | id),
  longitudinal_data,
  control = list(parallel = TRUE, verbose = 1)
)

# Level 2a: + random frequency only
mfit2a <- MarginalODE(
  observed ~ biomarker + velocity + drug + (biomarker | id),
  longitudinal_data,
  control = list(parallel = TRUE, verbose = 1)
)

# Level 2b: + random damping only
mfit2b <- MarginalODE(
  observed ~ biomarker + velocity + drug + (velocity | id),
  longitudinal_data,
  control = list(parallel = TRUE, verbose = 1)
)

# Level 2c: + random frequency & damping
mfit2c <- MarginalODE(
  observed ~ biomarker + velocity + drug + (biomarker + velocity | id),
  longitudinal_data,
  control = list(parallel = TRUE, verbose = 1)
)
```

```{r compare-marginal}
knitr::kable(data.frame(
  Model = c("Level 0: fixed dynamics",
            "Level 1: random forcing",
            "Level 2a: random frequency",
            "Level 2b: random damping",
            "Level 2c: random freq + damp"),
  RE = c("(none)", "(1|id)", "(biomarker|id)",
         "(velocity|id)", "(biomarker+velocity|id)"),
  n_RE = c(2, 3, 3, 3, 4),
  logLik = c(mfit0$logLik, mfit1$logLik, mfit2a$logLik,
             mfit2b$logLik, mfit2c$logLik),
  AIC = c(mfit0$AIC, mfit1$AIC, mfit2a$AIC,
          mfit2b$AIC, mfit2c$AIC),
  Converged = c(mfit0$convergence$converged,
                mfit1$convergence$converged,
                mfit2a$convergence$converged,
                mfit2b$convergence$converged,
                mfit2c$convergence$converged)
), digits = 1, caption = "MarginalODE model comparison")
```

## Diagnosing Level 2

When Level 2 converges, check the dynamics RE variances. Values near
zero indicate the data does not support that RE.

```{r re-variance}
re_sd <- sqrt(diag(mfit2c$random_effect_sigma))
names(re_sd) <- c("init_biomarker", "init_velocity",
                   "dyn_biomarker", "dyn_velocity")
knitr::kable(
  data.frame(RE = names(re_sd), SD = round(re_sd, 4)),
  caption = "Random effect standard deviations (Level 2)"
)
```

## Model Selection

```{r decision}
# Collect all candidates
candidates <- list(
  list(name = "Level 0", fit = mfit0,
       formula = observed ~ biomarker + velocity + drug),
  list(name = "Level 1", fit = mfit1,
       formula = observed ~ biomarker + velocity + drug + (1 | id)),
  list(name = "Level 2a", fit = mfit2a,
       formula = observed ~ biomarker + velocity + drug + (biomarker | id)),
  list(name = "Level 2b", fit = mfit2b,
       formula = observed ~ biomarker + velocity + drug + (velocity | id)),
  list(name = "Level 2c", fit = mfit2c,
       formula = observed ~ biomarker + velocity + drug +
         (biomarker + velocity | id))
)

# Select: best AIC among converged models
converged <- Filter(function(x) x$fit$convergence$converged, candidates)
if (length(converged) == 0) {
  # Fallback: use Level 1 even if not converged
  best <- candidates[[2]]
} else {
  aics <- sapply(converged, function(x) x$fit$AIC)
  best <- converged[[which.min(aics)]]
}
cat(sprintf("Selected: %s (AIC = %.1f)\n", best$name, best$fit$AIC))
selected_formula <- best$formula
```

Non-convergence at Level 2 is not a software failure; it signals that
the data cannot identify subject-specific dynamics. For PBC albumin,
trajectories are mostly monotone with little curvature variation across
patients, so a simpler model is the appropriate choice.

# Joint Model

```{r fit-joint}
fit <- JointODE(
  longitudinal_formula = selected_formula,
  survival_formula = Surv(time, status) ~ drug + age + sex,
  longitudinal_data = longitudinal_data,
  survival_data = survival_data,
  init = "marginal",
  control = list(parallel = TRUE, verbose = 1)
)
summary(fit)
```

## Interpreting ODE Parameters

```{r ode-interpretation}
cf <- coef(fit)
b1 <- cf["longitudinal:biomarker"]
b2 <- cf["longitudinal:velocity"]

if (b1 < 0) {
  omega <- sqrt(-b1)
  xi <- -b2 / (2 * omega)
  cat(sprintf("Period: %.1f years | Damping ratio: %.3f\n",
              2 * pi / omega, xi))
  cat(if (xi > 1) "Overdamped: monotonic decay.\n"
      else if (xi > 0) "Underdamped: damped oscillation.\n"
      else "Unstable: negative damping.\n")
}
```

A period much longer than follow-up (here ~28 years vs. ~15 years of
follow-up) means the ODE effectively describes exponential decay rather
than oscillation. This is consistent with the biology of albumin
decline in PBC.

## Survival Association

The hazard links both biomarker level and velocity to risk:

$$\log h_i(t) = \alpha_1 m_i(t) + \alpha_2 m_i'(t) + W_i'\beta + \text{baseline}$$

Unlike standard joint models where `slope()` must be explicitly
specified, JointODE derives velocity naturally from the ODE solution.

# Comparison with Alternative Approaches

## Extended Cox Model

A time-dependent Cox model treats the biomarker as a step-function
covariate updated at each observation time. This is a common baseline
that ignores measurement error and interpolation.

```{r fit-cox-td}
# Build counting-process dataset via tmerge
base <- survival_data |> select(id, time, status, drug, age, sex)
td <- tmerge(base, base, id = id, event = event(time, status))
td <- tmerge(td, longitudinal_data, id = id,
             observed = tdc(time, observed))

cox_td <- coxph(
  Surv(tstart, tstop, event) ~ observed + drug + age + sex,
  data = td
)
```

## JMbayes2

```{r fit-jmbayes2}
cox_fit <- coxph(
  Surv(time, status) ~ drug + age + sex,
  data = survival_data, x = TRUE
)
lme_fit <- lme(
  observed ~ ns(time, 3) * drug,
  random = ~ time | id,
  data = longitudinal_data
)

jm_val <- jm(cox_fit, lme_fit, time_var = "time",
             functional_forms = list("observed" = ~ value(observed)))
jm_both <- jm(cox_fit, lme_fit, time_var = "time",
              functional_forms = list(
                "observed" = ~ value(observed) + slope(observed)
              ))
```

## Hazard Coefficient Comparison

```{r comparison-table}
ode_cf <- coef(fit)
ode_se <- sqrt(diag(vcov(fit)))
cox_s <- summary(cox_td)$coefficients
jv <- summary(jm_val)$Survival
jb <- summary(jm_both)$Survival

f <- function(est, se) sprintf("%.3f (%.3f)", est, se)

knitr::kable(data.frame(
  Parameter = c("Biomarker level", "Biomarker velocity",
                "Drug", "Age", "Sex"),
  JointODE = c(
    f(ode_cf["hazard:alpha_1"], ode_se["hazard:alpha_1"]),
    f(ode_cf["hazard:alpha_2"], ode_se["hazard:alpha_2"]),
    f(ode_cf["hazard:drug"], ode_se["hazard:drug"]),
    f(ode_cf["hazard:age"], ode_se["hazard:age"]),
    f(ode_cf["hazard:sex"], ode_se["hazard:sex"])),
  Ext.Cox = c(
    f(cox_s["observed", 1], cox_s["observed", 3]), "--",
    f(cox_s["drug", 1], cox_s["drug", 3]),
    f(cox_s["age", 1], cox_s["age", 3]),
    f(cox_s["sex", 1], cox_s["sex", 3])),
  JM.value = c(
    f(jv["value(observed)", 1], jv["value(observed)", 2]), "--",
    f(jv["drug", 1], jv["drug", 2]),
    f(jv["age", 1], jv["age", 2]),
    f(jv["sex", 1], jv["sex", 2])),
  JM.value.slope = c(
    f(jb["value(observed)", 1], jb["value(observed)", 2]),
    f(jb["slope(observed)", 1], jb["slope(observed)", 2]),
    f(jb["drug", 1], jb["drug", 2]),
    f(jb["age", 1], jb["age", 2]),
    f(jb["sex", 1], jb["sex", 2]))
), align = c("l", "r", "r", "r", "r"),
caption = "Hazard coefficients: estimate (SE)")
```

The extended Cox model cannot estimate a velocity effect and is
susceptible to bias from measurement error. JMbayes2 requires
explicitly specifying `slope()` as a functional form, while JointODE
provides both value and velocity associations by construction.

# Diagnostics

```{r predicted-trajectories, fig.height=6}
pred <- predict(fit)

pred |>
  mutate(id = as.integer(id)) |>
  filter(id %in% sample_ids) |>
  left_join(
    longitudinal_data |> select(id, time, observed),
    by = c("id", "time")
  ) |>
  ggplot(aes(x = time)) +
  geom_point(aes(y = observed), alpha = 0.4, size = 1) +
  geom_line(aes(y = biomarker), color = "steelblue") +
  facet_wrap(~id, scales = "free_y") +
  labs(x = "Time (years)", y = "Log Albumin (standardized)") +
  theme_minimal(base_size = 10)
```

```{r c-index}
cat(sprintf("C-index: %.3f\n", fit$cindex))
```

# Workflow Summary

1. **Standardize** the response
2. **Fit MarginalODE** at increasing complexity (Level 0 / 1 / 2)
3. **Select RE** based on convergence, AIC, and RE variance diagnostics
4. **Fit JointODE** with `init = "marginal"`
5. **Compare** with standard approaches (JMbayes2)

```{r session-info}
sessionInfo()
```
