---
title: "JointODE Examples: PBC Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{JointODE Examples: PBC Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  eval = requireNamespace("JMbayes2", quietly = TRUE),
  fig.width = 8, fig.height = 5
)
```

This vignette demonstrates a complete JointODE workflow using PBC
(Primary Biliary Cirrhosis) serum albumin data from the JMbayes2
package. The workflow has four steps:

1. Prepare and standardize data
2. Select random effects structure via `MarginalODE`
3. Fit the joint model via `JointODE`
4. Interpret and compare results

# Step 1: Data Preparation

```{r data-prep}
library(JointODE)
library(JMbayes2)
library(survival)
library(dplyr)
library(ggplot2)

survival_data <- pbc2.id |>
  transmute(
    id = as.integer(id), time = years, status = status2,
    drug = as.numeric(drug == "D-penicil"),
    age = scale(age)[, 1],
    sex = as.numeric(sex == "female")
  )

longitudinal_data <- pbc2 |>
  transmute(id = as.integer(id), time = year, observed = log(albumin)) |>
  left_join(survival_data |> select(id, drug, age, sex), by = "id")

# Standardize: ODE parameters scale with the response
alb_mean <- mean(longitudinal_data$observed)
alb_sd <- sd(longitudinal_data$observed)
longitudinal_data$observed <- (longitudinal_data$observed - alb_mean) / alb_sd

cat(sprintf(
  "Subjects: %d | Events: %d (%.0f%%) | Observations: %d\n",
  nrow(survival_data), sum(survival_data$status),
  100 * mean(survival_data$status), nrow(longitudinal_data)
))
```

```{r trajectories, fig.height=6}
set.seed(42)
sample_ids <- sample(unique(longitudinal_data$id), 20)

longitudinal_data |>
  filter(id %in% sample_ids) |>
  left_join(survival_data |> select(id, status), by = "id") |>
  ggplot(aes(time, observed, group = id, color = factor(status))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1, alpha = 0.6) +
  scale_color_manual(
    values = c("0" = "steelblue", "1" = "tomato"),
    labels = c("Censored", "Event"), name = "Status"
  ) +
  labs(x = "Time (years)", y = "Log Albumin (standardized)") +
  theme_minimal(base_size = 12)
```

# Step 2: Random Effects Selection with MarginalODE

The ODE is $m''(t) + 2\xi\omega\, m'(t) + \omega^2 m(t) = f(t)$,
with initial conditions $m(0) = m_0$, $m'(0) = v_0$. The formula uses
two reserved keywords:

- `omega`: latent log-frequency parameter; physical $\omega_i=\exp(\text{omega}_i)$
- `xi`: latent damping root; physical $\xi_i=\text{xi}_i^2$

Random effects on initial states $(m_{0,i}, v_{0,i})$ are always
included. We fit models at increasing complexity to find the right
RE structure.

```{r marginal-fits}
ctrl <- list(parallel = TRUE, verbose = 1)

# Level 0: fixed dynamics, random initial state only
mfit0 <- MarginalODE(
  observed ~ omega + xi + drug,
  longitudinal_data,
  control = ctrl
)

# Level 1: + random forcing intercept
mfit1 <- MarginalODE(
  observed ~ omega + xi + drug + (1 | id),
  longitudinal_data,
  control = ctrl
)

# Level 2a: + random frequency
mfit2a <- MarginalODE(
  observed ~ omega + xi + drug + (omega | id),
  longitudinal_data,
  control = ctrl
)

# Level 2b: + random damping
mfit2b <- MarginalODE(
  observed ~ omega + xi + drug + (xi | id),
  longitudinal_data,
  control = ctrl
)

# Level 2c: + random frequency & damping
mfit2c <- MarginalODE(
  observed ~ omega + xi + drug + (omega + xi | id),
  longitudinal_data,
  control = ctrl
)
```

```{r compare-marginal}
fits <- list(mfit0, mfit1, mfit2a, mfit2b, mfit2c)
knitr::kable(data.frame(
  Model = c(
    "Level 0: fixed dynamics",
    "Level 1: + random forcing",
    "Level 2a: + random frequency",
    "Level 2b: + random damping",
    "Level 2c: + random freq & damp"
  ),
  RE = c(
    "(none)", "(1|id)", "(biomarker|id)",
    "(velocity|id)", "(biomarker+velocity|id)"
  ),
  n_RE = sapply(fits, function(f) ncol(f$random_effects)),
  logLik = sapply(fits, `[[`, "logLik"),
  AIC = sapply(fits, `[[`, "AIC"),
  Converged = sapply(fits, function(f) f$convergence$converged)
), digits = 1, caption = "MarginalODE model comparison")
```

## Selecting the Best Model

```{r decision}
converged_idx <- which(sapply(fits, function(f) f$convergence$converged))
aics <- sapply(fits[converged_idx], `[[`, "AIC")
best_idx <- converged_idx[which.min(aics)]
best_marginal <- fits[[best_idx]]
cat(sprintf(
  "Selected: Level %s (AIC = %.1f)\n",
  c("0", "1", "2a", "2b", "2c")[best_idx], best_marginal$AIC
))
```

```{r best-summary}
summary(best_marginal)
```

The best model (Level 2b) adds random damping. The large damping
ratio ($\xi \gg 1$) indicates overdamped dynamics: albumin declines
monotonically rather than oscillating, consistent with PBC biology.

## RE Variance Diagnostics

```{r re-variance}
re_sigma <- best_marginal$random_effect_sigma
knitr::kable(
  data.frame(RE = rownames(re_sigma), SD = round(sqrt(diag(re_sigma)), 4)),
  caption = "Random effect standard deviations"
)
```

# Step 3: Joint Model

We now link the longitudinal trajectories to survival using `JointODE`.
The `init = "marginal"` option warm-starts from the MarginalODE fit.

```{r fit-joint}
# Use the formula from the best marginal model
selected_formula <- list(
  observed ~ omega + xi + drug,
  observed ~ omega + xi + drug + (1 | id),
  observed ~ omega + xi + drug + (omega | id),
  observed ~ omega + xi + drug + (xi | id),
  observed ~ omega + xi + drug + (omega + xi | id)
)[[best_idx]]

fit <- JointODE(
  longitudinal_formula = selected_formula,
  survival_formula = Surv(time, status) ~ drug + age + sex,
  longitudinal_data = longitudinal_data,
  survival_data = survival_data,
  init = "marginal",
  control = list(parallel = TRUE, verbose = 1)
)
summary(fit)
```

# Step 4: Interpretation

## ODE Dynamics

The summary reports $\omega$ (natural frequency) and $\xi$ (damping
ratio) with delta-method standard errors:

```{r ode-interpretation}
summary(fit)$derived_params
```

## Survival Association

The hazard model links biomarker value and velocity to event risk:

$$\log h_i(t) = \alpha_\text{value}\, m_i(t) + \alpha_\text{velocity}\, m_i'(t) + W_i'\beta + \text{baseline}$$

JointODE provides both associations by construction, without needing
to specify `slope()` as in JMbayes2.

## Predicted Trajectories

```{r predicted-trajectories, fig.height=6}
pred <- predict(fit)

pred |>
  mutate(id = as.integer(id)) |>
  filter(id %in% sample_ids) |>
  left_join(
    longitudinal_data |> select(id, time, observed),
    by = c("id", "time")
  ) |>
  ggplot(aes(x = time)) +
  geom_point(aes(y = observed), alpha = 0.4, size = 1) +
  geom_line(aes(y = biomarker), color = "steelblue") +
  facet_wrap(~id, scales = "free_y") +
  labs(x = "Time (years)", y = "Log Albumin (standardized)") +
  theme_minimal(base_size = 10)
```

```{r c-index}
cat(sprintf("C-index: %.3f\n", fit$cindex))
```

# Comparison with Alternative Approaches

## Extended Cox Model

```{r fit-cox-td}
base <- survival_data |> select(id, time, status, drug, age, sex)
td <- tmerge(base, base, id = id, event = event(time, status))
td <- tmerge(td, longitudinal_data,
  id = id,
  observed = tdc(time, observed)
)

cox_td <- coxph(
  Surv(tstart, tstop, event) ~ observed + drug + age + sex,
  data = td
)
```

## JMbayes2

```{r fit-jmbayes2}
cox_fit <- coxph(
  Surv(time, status) ~ drug + age + sex,
  data = survival_data, x = TRUE
)
lme_fit <- lme(
  observed ~ ns(time, 3) * drug,
  random = ~ time | id,
  data = longitudinal_data
)

jm_val <- jm(cox_fit, lme_fit,
  time_var = "time",
  functional_forms = list("observed" = ~ value(observed))
)
jm_both <- jm(cox_fit, lme_fit,
  time_var = "time",
  functional_forms = list(
    "observed" = ~ value(observed) + slope(observed)
  )
)
```

## Hazard Coefficient Comparison

```{r comparison-table}
ode_cf <- coef(fit)
ode_se <- sqrt(diag(vcov(fit)))
cox_s <- summary(cox_td)$coefficients
jv <- summary(jm_val)$Survival
jb <- summary(jm_both)$Survival

f <- function(est, se) sprintf("%.3f (%.3f)", est, se)

knitr::kable(
  data.frame(
    Parameter = c(
      "Biomarker level", "Biomarker velocity",
      "Drug", "Age", "Sex"
    ),
    JointODE = c(
      f(ode_cf["hazard:value"], ode_se["hazard:value"]),
      f(ode_cf["hazard:velocity"], ode_se["hazard:velocity"]),
      f(ode_cf["hazard:drug"], ode_se["hazard:drug"]),
      f(ode_cf["hazard:age"], ode_se["hazard:age"]),
      f(ode_cf["hazard:sex"], ode_se["hazard:sex"])
    ),
    Ext.Cox = c(
      f(cox_s["observed", 1], cox_s["observed", 3]), "--",
      f(cox_s["drug", 1], cox_s["drug", 3]),
      f(cox_s["age", 1], cox_s["age", 3]),
      f(cox_s["sex", 1], cox_s["sex", 3])
    ),
    JM.value = c(
      f(jv["value(observed)", 1], jv["value(observed)", 2]), "--",
      f(jv["drug", 1], jv["drug", 2]),
      f(jv["age", 1], jv["age", 2]),
      f(jv["sex", 1], jv["sex", 2])
    ),
    JM.value.slope = c(
      f(jb["value(observed)", 1], jb["value(observed)", 2]),
      f(jb["slope(observed)", 1], jb["slope(observed)", 2]),
      f(jb["drug", 1], jb["drug", 2]),
      f(jb["age", 1], jb["age", 2]),
      f(jb["sex", 1], jb["sex", 2])
    )
  ),
  align = c("l", "r", "r", "r", "r"),
  caption = "Hazard coefficients: estimate (SE)"
)
```

The extended Cox model cannot estimate a velocity effect and is
susceptible to bias from measurement error. JMbayes2 requires
explicitly specifying `slope()` as a functional form, while JointODE
provides both associations by construction.

```{r session-info}
sessionInfo()
```
