---
title: "JointODE Examples: PBC Data Analysis"
author: "JointODE Package"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  cache = FALSE
)
```

# Introduction

This document demonstrates a complete analysis workflow using the **JointODE** package with the PBC (Primary Biliary Cirrhosis) dataset. We will:

1. Prepare and explore the data
2. Fit a JointODE model
3. Compare with standard joint models (JMbayes2)
4. Generate predictions and visualizations

# Data Preparation

## Load Packages

```{r load-packages}
library(JointODE)
library(survival)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
```

## Load and Process PBC Data

The PBC data contains information from a Mayo Clinic trial on primary biliary cirrhosis (PBC) of the liver conducted between 1974 and 1984.

```{r load-data}
# Load PBC data from survival package
# pbc: baseline and survival data (one row per patient)
# pbcseq: longitudinal measurements (multiple rows per patient)
data(pbc, package = "survival")
data(pbcseq, package = "survival")

# Prepare survival data (one row per patient)
survival_data <- pbc %>%
  filter(!is.na(trt)) %>%
  transmute(
    id = as.integer(id),
    time = time / 365.25,  # Convert days to years
    status = as.integer(status == 2),  # 2 = death, 1 = transplant, 0 = censored
    trt = as.numeric(trt == 2),  # Treatment indicator (1 = D-penicillamine)
    age = age,
    sex = as.numeric(sex == "f")  # Female indicator
  )

# Prepare longitudinal data from pbcseq
# Using log serum bilirubin as the biomarker
longitudinal_data <- pbcseq %>%
  filter(id %in% survival_data$id, !is.na(bili)) %>%
  transmute(
    id = as.integer(id),
    time = day / 365.25,  # Convert days to years
    observed = log(bili)  # Log serum bilirubin
  ) %>%
  # Add baseline covariates from survival data
  left_join(
    survival_data %>% select(id, trt, age, sex),
    by = "id"
  ) %>%
  arrange(id, time)
```

## Data Summary

```{r data-summary}
cat("Survival Data:\n")
cat(sprintf("  Patients: %d\n", nrow(survival_data)))
cat(sprintf("  Events: %d (%.1f%%)\n",
            sum(survival_data$status),
            100 * mean(survival_data$status)))
cat(sprintf("  Median follow-up: %.2f years\n",
            median(survival_data$time)))

cat("\nLongitudinal Data:\n")
cat(sprintf("  Total observations: %d\n", nrow(longitudinal_data)))
cat(sprintf("  Patients: %d\n", length(unique(longitudinal_data$id))))
cat(sprintf("  Avg measurements per patient: %.1f\n",
            nrow(longitudinal_data) / length(unique(longitudinal_data$id))))

cat("\nLog Serum Bilirubin Summary:\n")
print(summary(longitudinal_data$observed))
```

# Exploratory Data Analysis

## Individual Trajectories

```{r trajectories, fig.height=6}
# Plot sample of individual trajectories
set.seed(123)
sample_ids <- sample(unique(longitudinal_data$id), 20)

longitudinal_data %>%
  filter(id %in% sample_ids) %>%
  left_join(
    survival_data %>% select(id, status),
    by = "id"
  ) %>%
  ggplot(aes(x = time, y = observed, group = id, color = factor(status))) +
  geom_line(alpha = 0.6) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(
    values = c("0" = "#3498db", "1" = "#e74c3c"),
    labels = c("Censored", "Event"),
    name = "Status"
  ) +
  labs(
    title = "Log Serum Bilirubin Trajectories (Sample of 20 Patients)",
    x = "Time (years)",
    y = "Log Serum Bilirubin"
  ) +
  theme_minimal(base_size = 12)
```

**Observation:** Patients who experienced events (red) tend to have higher and increasing bilirubin levels compared to censored patients (blue).

## Mean Trajectories by Event Status

```{r mean-trajectories}
longitudinal_data %>%
  left_join(
    survival_data %>% select(id, status),
    by = "id"
  ) %>%
  group_by(status, time_bin = cut(time, breaks = seq(0, 12, by = 0.5))) %>%
  summarise(
    mean_bili = mean(observed),
    se_bili = sd(observed) / sqrt(n()),
    time_mid = mean(time),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = time_mid, y = mean_bili, color = factor(status))) +
  geom_line(linewidth = 1) +
  geom_ribbon(
    aes(ymin = mean_bili - 1.96 * se_bili,
        ymax = mean_bili + 1.96 * se_bili,
        fill = factor(status)),
    alpha = 0.2,
    color = NA
  ) +
  scale_color_manual(
    values = c("0" = "#3498db", "1" = "#e74c3c"),
    labels = c("Censored", "Event"),
    name = "Status"
  ) +
  scale_fill_manual(
    values = c("0" = "#3498db", "1" = "#e74c3c"),
    labels = c("Censored", "Event"),
    name = "Status"
  ) +
  labs(
    title = "Mean Log Serum Bilirubin by Event Status",
    x = "Time (years)",
    y = "Mean Log Serum Bilirubin",
    subtitle = "Shaded area represents 95% confidence interval"
  ) +
  theme_minimal(base_size = 12)
```

**Observation:** Clear separation between event and censored groups, with diverging trajectories over time.

# Model Fitting

## JointODE Model

We fit a joint ODE model with:

- **Longitudinal submodel:** Log bilirubin modeled with both current value and rate of change
- **Survival submodel:** Cox model with treatment, age, and sex
- **Association:** Both biomarker value and velocity linked to hazard

```{r fit-jointode}
fit_ode <- JointODE(
  longitudinal_formula = observed ~ biomarker + velocity + trt +
    (biomarker + velocity | id),
  survival_formula = Surv(time, status) ~ trt + age + sex,
  longitudinal_data = longitudinal_data,
  survival_data = survival_data,
  control = list(verbose = 3)
)
```

### Model Summary

```{r summary-jointode}
summary(fit_ode)
```

### Extract Key Results

```{r results-jointode}
# Coefficients
coef_table <- as.data.frame(coef(fit_ode))
kable(coef_table, digits = 4, caption = "JointODE Model Coefficients")

# Model fit statistics
cat("\nModel Fit Statistics:\n")
cat(sprintf("  Log-likelihood: %.2f\n", logLik(fit_ode)))
cat(sprintf("  AIC: %.2f\n", AIC(fit_ode)))
cat(sprintf("  BIC: %.2f\n", BIC(fit_ode)))
cat(sprintf("  C-index: %.3f\n", fit_ode$cindex))
```

## Standard Joint Model (JMbayes2)

For comparison, we fit a standard joint model using JMbayes2.

```{r fit-jmbayes2}
if (requireNamespace("JMbayes2", quietly = TRUE)) {
  library(JMbayes2)
  library(nlme)

  # Linear mixed model
  lme_fit <- lme(
    observed ~ time + trt,
    random = ~ time | id,
    data = longitudinal_data,
    control = lmeControl(opt = "optim")
  )

  # Cox model
  cox_fit <- coxph(
    Surv(time, status) ~ trt + age + sex,
    data = survival_data,
    x = TRUE
  )

  # Joint model
  system.time({
    fit_jm <- jm(
      cox_fit,
      lme_fit,
      time_var = "time",
      functional_forms = ~ value(observed) + slope(observed)
    )
  })

  cat("\nJMbayes2 Model Summary:\n")
  print(summary(fit_jm))

  has_jm <- TRUE
} else {
  cat("JMbayes2 not installed. Install with: install.packages('JMbayes2')\n")
  has_jm <- FALSE
}
```

# Model Comparison

```{r model-comparison}
# Create comparison table
comparison <- data.frame(
  Model = "JointODE",
  LogLik = as.numeric(logLik(fit_ode)),
  AIC = AIC(fit_ode),
  BIC = BIC(fit_ode),
  Cindex = fit_ode$cindex
)

if (exists("fit_jm")) {
  comparison <- rbind(
    comparison,
    data.frame(
      Model = "JMbayes2",
      LogLik = as.numeric(logLik(fit_jm)),
      AIC = AIC(fit_jm),
      BIC = BIC(fit_jm),
      Cindex = NA
    )
  )

  comparison$Delta_AIC <- comparison$AIC - min(comparison$AIC)
  comparison$Delta_BIC <- comparison$BIC - min(comparison$BIC)
}

kable(comparison, digits = 2, caption = "Model Comparison")
```

**Interpretation:**

- Lower AIC/BIC indicates better model fit
- C-index measures discrimination ability (higher is better)
- Delta AIC/BIC > 10 suggests strong evidence for the better model

# Predictions

## Individual Patient Predictions

We select three patients to demonstrate prediction capabilities.

```{r predictions, fig.height=5}
# Select patients
set.seed(456)
predict_ids <- sample(unique(longitudinal_data$id), 3)

for (pid in predict_ids) {
  cat(sprintf("\n### Patient %d\n\n", pid))

  # Get patient data
  patient_long <- longitudinal_data %>% filter(id == pid)
  patient_surv <- survival_data %>% filter(id == pid)

  cat(sprintf("- Measurements: %d\n", nrow(patient_long)))
  cat(sprintf("- Follow-up: %.2f years\n", patient_surv$time))
  cat(sprintf("- Event: %s\n\n", ifelse(patient_surv$status == 1, "Yes", "No")))

  # Predict biomarker trajectory
  pred_times <- seq(0, max(patient_long$time) + 2, by = 0.1)

  pred_ode <- predict(
    fit_ode,
    newdata = list(
      longitudinal = patient_long,
      survival = patient_surv
    ),
    times = pred_times,
    return_data = TRUE
  )

  # Plot
  p <- ggplot() +
    geom_line(
      data = pred_ode,
      aes(x = time, y = biomarker_mean),
      color = "#3498db",
      linewidth = 1.2
    ) +
    geom_ribbon(
      data = pred_ode,
      aes(x = time,
          ymin = biomarker_mean - 1.96 * biomarker_se,
          ymax = biomarker_mean + 1.96 * biomarker_se),
      fill = "#3498db",
      alpha = 0.2
    ) +
    geom_point(
      data = patient_long,
      aes(x = time, y = observed),
      color = "#2c3e50",
      size = 3
    ) +
    geom_vline(
      xintercept = max(patient_long$time),
      linetype = "dashed",
      color = "#7f8c8d"
    ) +
    annotate(
      "text",
      x = max(patient_long$time),
      y = max(pred_ode$biomarker_mean),
      label = "Last observation",
      hjust = -0.1,
      color = "#7f8c8d"
    ) +
    labs(
      title = sprintf("Patient %d: Log Serum Bilirubin Prediction", pid),
      x = "Time (years)",
      y = "Log Serum Bilirubin",
      subtitle = "Observed (points), Predicted (line), 95% CI (ribbon)"
    ) +
    theme_minimal(base_size = 12)

  print(p)
  cat("\n\n")
}
```

## Dynamic Survival Predictions

For censored patients, we can predict future survival probabilities.

```{r survival-predictions}
# Select a censored patient
censored_ids <- survival_data %>%
  filter(status == 0) %>%
  pull(id)

if (length(censored_ids) > 0) {
  example_id <- censored_ids[1]

  patient_long <- longitudinal_data %>% filter(id == example_id)
  patient_surv <- survival_data %>% filter(id == example_id)

  cat(sprintf("Patient %d (Censored)\n", example_id))
  cat(sprintf("Last observation time: %.2f years\n\n", patient_surv$time))

  # Predict survival at future time points
  future_times <- seq(patient_surv$time, patient_surv$time + 5, by = 0.5)

  pred_surv <- predict(
    fit_ode,
    newdata = list(
      longitudinal = patient_long,
      survival = patient_surv
    ),
    times = future_times,
    type = "survival"
  )

  kable(
    pred_surv,
    digits = 3,
    caption = sprintf("Survival Predictions for Patient %d", example_id)
  )
}
```

# Conclusion

This analysis demonstrates:

1. **JointODE modeling** of longitudinal biomarkers and survival outcomes
2. **Comparison** with standard joint models
3. **Individual predictions** for future biomarker values and survival probabilities

The JointODE framework allows flexible modeling of complex biomarker dynamics through ODE systems, which can capture non-linear patterns and rate-of-change effects.

# Session Info

```{r session-info}
sessionInfo()
```
