---
title: "JointODE Examples: PBC Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{JointODE Examples: PBC Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  eval = requireNamespace("JMbayes2", quietly = TRUE),
  fig.width = 8, fig.height = 5
)
```

This vignette demonstrates a complete JointODE workflow using PBC
(Primary Biliary Cirrhosis) serum albumin data from the JMbayes2
package. The goal is to show the modeling sequence rather than to claim a
single definitive clinical model. The workflow has four steps:

1. Prepare longitudinal and survival data
2. Select a stable longitudinal random-effects structure with `MarginalODE`
3. Fit the joint model with the selected longitudinal structure
4. Compare the fitted association estimates with common alternatives

# Step 1: Data Preparation

```{r data-prep}
library(JointODE)
library(JMbayes2)
library(survival)
library(dplyr)
library(ggplot2)
library(splines)

survival_data <- pbc2.id |>
  transmute(
    id = as.integer(id), time = years, status = status2,
    drug = as.numeric(drug == "D-penicil"),
    age = scale(age)[, 1],
    sex = as.numeric(sex == "female")
  )

longitudinal_data <- pbc2 |>
  transmute(id = as.integer(id), time = year, observed = log(albumin)) |>
  left_join(survival_data |> select(id, drug, age, sex), by = "id")

# Standardize the response; the input time scale remains the observed years.
alb_mean <- mean(longitudinal_data$observed)
alb_sd <- sd(longitudinal_data$observed)
longitudinal_data$observed <- (longitudinal_data$observed - alb_mean) / alb_sd

cat(sprintf(
  "Subjects: %d | Events: %d (%.0f%%) | Observations: %d\n",
  nrow(survival_data), sum(survival_data$status),
  100 * mean(survival_data$status), nrow(longitudinal_data)
))
```

```{r trajectories, fig.height=6}
set.seed(42)
sample_ids <- sample(unique(longitudinal_data$id), 20)

longitudinal_data |>
  filter(id %in% sample_ids) |>
  left_join(survival_data |> select(id, status), by = "id") |>
  ggplot(aes(time, observed, group = id, color = factor(status))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1, alpha = 0.6) +
  scale_color_manual(
    values = c("0" = "steelblue", "1" = "tomato"),
    labels = c("Censored", "Event"), name = "Status"
  ) +
  labs(x = "Time (years)", y = "Log Albumin (standardized)") +
  theme_minimal(base_size = 12)
```

# Step 2: Random Effects Selection with MarginalODE

The ODE is $m''(t) + 2\xi\omega\, m'(t) + \omega^2 m(t) = f(t)$,
with initial conditions $m(0)=m_0$ and $m'(0)=v_0$. The formula uses two reserved
keywords:

- `biomarker`: the latent $\log \omega_i^2$ parameter
- `velocity`: the latent $\log(2\xi_i\omega_i)$ parameter

Random effects on the initial biomarker level and initial velocity are always
included internally. Formula random effects add heterogeneity in the forcing
function or in the ODE parameters. As in `lme4`, `|` requests a correlated
covariance block and `||` requests independent random effects.

For real data, a practical first pass is to start with independent random
effects on forcing, frequency, and damping. If that model is not numerically
stable, remove one dynamic random effect at a time before falling back to a
random forcing intercept only.

```{r marginal-fits}
ctrl <- list(parallel = TRUE, verbose = 1)

# Full diagonal: random forcing, frequency, and damping
mfit_full <- MarginalODE(
  observed ~ biomarker + velocity + drug + (1 + biomarker + velocity || id),
  longitudinal_data,
  control = ctrl
)

# Drop random damping
mfit_no_velocity <- MarginalODE(
  observed ~ biomarker + velocity + drug + (1 + biomarker || id),
  longitudinal_data,
  control = ctrl
)

# Drop random frequency
mfit_no_biomarker <- MarginalODE(
  observed ~ biomarker + velocity + drug + (1 + velocity || id),
  longitudinal_data,
  control = ctrl
)

# Random forcing only
mfit_forcing <- MarginalODE(
  observed ~ biomarker + velocity + drug + (1 || id),
  longitudinal_data,
  control = ctrl
)
```

```{r compare-marginal}
fits <- list(mfit_full, mfit_no_velocity, mfit_no_biomarker, mfit_forcing)
knitr::kable(data.frame(
  Model = c(
    "Full diagonal",
    "No random damping",
    "No random frequency",
    "Random forcing only"
  ),
  RE = c(
    "(1+biomarker+velocity||id)",
    "(1+biomarker||id)",
    "(1+velocity||id)",
    "(1||id)"
  ),
  n_RE = sapply(fits, function(f) ncol(f$random_effects)),
  logLik = sapply(fits, `[[`, "logLik"),
  AIC = sapply(fits, `[[`, "AIC"),
  Converged = sapply(fits, function(f) f$convergence$converged)
), digits = 1, caption = "MarginalODE model comparison")
```

## Selecting the Best Model

```{r decision}
converged_idx <- which(sapply(fits, function(f) f$convergence$converged))
best_idx <- converged_idx[1]
best_marginal <- fits[[best_idx]]
cat(sprintf(
  "Selected: %s (AIC = %.1f)\n",
  c("full diagonal", "no random damping", "no random frequency",
    "random forcing only")[best_idx],
  best_marginal$AIC
))
```

```{r best-summary}
summary(best_marginal)
```

The selected model is the first cleanly converged model along the prespecified
simplification path. A large damping ratio ($\xi \gg 1$), when present,
indicates overdamped dynamics: albumin changes monotonically rather than
oscillating.

## RE Variance Diagnostics

```{r re-variance}
re_sigma <- best_marginal$random_effect_sigma
knitr::kable(
  data.frame(RE = rownames(re_sigma), SD = round(sqrt(diag(re_sigma)), 4)),
  caption = "Random effect standard deviations"
)
```

# Step 3: Joint Model

We now link the longitudinal trajectories to survival using `JointODE`.
The `init = "marginal"` option warm-starts from the MarginalODE fit.

```{r fit-joint}
# Use the formula from the best marginal model
selected_formula <- list(
  observed ~ biomarker + velocity + drug + (1 + biomarker + velocity || id),
  observed ~ biomarker + velocity + drug + (1 + biomarker || id),
  observed ~ biomarker + velocity + drug + (1 + velocity || id),
  observed ~ biomarker + velocity + drug + (1 || id)
)[[best_idx]]

fit <- JointODE(
  longitudinal_formula = selected_formula,
  survival_formula = Surv(time, status) ~ drug + age + sex,
  longitudinal_data = longitudinal_data,
  survival_data = survival_data,
  init = "marginal",
  control = list(parallel = TRUE, verbose = 1)
)
summary(fit)
```

# Step 4: Interpretation

## ODE Dynamics

The summary reports $\omega$ (natural frequency) and $\xi$ (damping
ratio) with delta-method standard errors:

```{r ode-interpretation}
summary(fit)$derived_params
```

## Survival Association

The hazard model links biomarker value and velocity to event risk:

$$\log h_i(t) = \alpha_\text{value}\, m_i(t) + \alpha_\text{velocity}\, m_i'(t) + W_i'\beta + \text{baseline}$$

JointODE provides both associations by construction, without needing
to specify `slope()` as in JMbayes2.

## Predicted Trajectories

```{r predicted-trajectories, fig.height=6}
pred <- predict(fit)

pred |>
  mutate(id = as.integer(id)) |>
  filter(id %in% sample_ids) |>
  left_join(
    longitudinal_data |> select(id, time, observed),
    by = c("id", "time")
  ) |>
  ggplot(aes(x = time)) +
  geom_point(aes(y = observed), alpha = 0.4, size = 1) +
  geom_line(aes(y = biomarker), color = "steelblue") +
  facet_wrap(~id, scales = "free_y") +
  labs(x = "Time (years)", y = "Log Albumin (standardized)") +
  theme_minimal(base_size = 10)
```

```{r c-index}
cat(sprintf("C-index: %.3f\n", fit$cindex))
```

# Comparison with Alternative Approaches

## Extended Cox Model

```{r fit-cox-td}
base <- survival_data |> select(id, time, status, drug, age, sex)
td <- tmerge(base, base, id = id, event = event(time, status))
td <- tmerge(td, longitudinal_data,
  id = id,
  observed = tdc(time, observed)
)

cox_td <- coxph(
  Surv(tstart, tstop, event) ~ observed + drug + age + sex,
  data = td
)
```

## JMbayes2

```{r fit-jmbayes2}
cox_fit <- coxph(
  Surv(time, status) ~ drug + age + sex,
  data = survival_data, x = TRUE
)
lme_fit <- lme(
  observed ~ ns(time, 3) * drug,
  random = ~ time | id,
  data = longitudinal_data
)

jm_val <- jm(cox_fit, lme_fit,
  time_var = "time",
  functional_forms = list("observed" = ~ value(observed))
)
jm_both <- jm(cox_fit, lme_fit,
  time_var = "time",
  functional_forms = list(
    "observed" = ~ value(observed) + slope(observed)
  )
)
```

## Hazard Coefficient Comparison

```{r comparison-table}
ode_cf <- coef(fit)
ode_se <- sqrt(diag(vcov(fit)))
cox_s <- summary(cox_td)$coefficients
jv <- summary(jm_val)$Survival
jb <- summary(jm_both)$Survival

f <- function(est, se) sprintf("%.3f (%.3f)", est, se)

knitr::kable(
  data.frame(
    Parameter = c(
      "Biomarker level", "Biomarker velocity",
      "Drug", "Age", "Sex"
    ),
    JointODE = c(
      f(ode_cf["hazard:value"], ode_se["hazard:value"]),
      f(ode_cf["hazard:velocity"], ode_se["hazard:velocity"]),
      f(ode_cf["hazard:drug"], ode_se["hazard:drug"]),
      f(ode_cf["hazard:age"], ode_se["hazard:age"]),
      f(ode_cf["hazard:sex"], ode_se["hazard:sex"])
    ),
    Ext.Cox = c(
      f(cox_s["observed", 1], cox_s["observed", 3]), "--",
      f(cox_s["drug", 1], cox_s["drug", 3]),
      f(cox_s["age", 1], cox_s["age", 3]),
      f(cox_s["sex", 1], cox_s["sex", 3])
    ),
    JM.value = c(
      f(jv["value(observed)", 1], jv["value(observed)", 2]), "--",
      f(jv["drug", 1], jv["drug", 2]),
      f(jv["age", 1], jv["age", 2]),
      f(jv["sex", 1], jv["sex", 2])
    ),
    JM.value.slope = c(
      f(jb["value(observed)", 1], jb["value(observed)", 2]),
      f(jb["slope(observed)", 1], jb["slope(observed)", 2]),
      f(jb["drug", 1], jb["drug", 2]),
      f(jb["age", 1], jb["age", 2]),
      f(jb["sex", 1], jb["sex", 2])
    )
  ),
  align = c("l", "r", "r", "r", "r"),
  caption = "Hazard coefficients: estimate (SE)"
)
```

The extended Cox model cannot estimate a velocity effect and is
susceptible to bias from measurement error. JMbayes2 requires
explicitly specifying `slope()` as a functional form, while JointODE
provides both associations by construction.

```{r session-info}
sessionInfo()
```
