---
title: "JointODE Examples: PBC Data Analysis"
author: "JointODE Package"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Data Preparation

```{r load-packages}
library(dplyr)
library(ggplot2)
library(knitr)
library(splines)
library(nlme)
library(survival)
library(JMbayes2)
library(JointODE)
```

We use the PBC dataset bundled with JMbayes2. `pbc2` contains repeated longitudinal measurements and `pbc2.id` contains one row per patient with baseline covariates and survival outcome. Log serum bilirubin serves as the longitudinal biomarker.

```{r load-data}
# Survival data: one row per patient
survival_data <- pbc2.id %>%
  transmute(
    id = as.integer(id),
    time = years,
    status = status2,
    drug = as.numeric(drug == "D-penicil"),
    age = age,
    sex = as.numeric(sex == "female")
  )

# Longitudinal data: repeated measurements
longitudinal_data <- pbc2 %>%
  transmute(
    id = as.integer(id),
    time = year,
    observed = log(serBilir)
  ) %>%
  left_join(survival_data %>% select(id, drug, age, sex), by = "id")

# # Standardize for numerical stability
bili_mean <- mean(longitudinal_data$observed)
bili_sd <- sd(longitudinal_data$observed)
longitudinal_data$observed <- (longitudinal_data$observed - bili_mean) / bili_sd
```

```{r data-summary}
cat(sprintf("Patients: %d | Events: %d (%.0f%%) | Observations: %d | Avg per patient: %.1f\n",
            nrow(survival_data), sum(survival_data$status),
            100 * mean(survival_data$status), nrow(longitudinal_data),
            nrow(longitudinal_data) / nrow(survival_data)))
```

# Exploratory Data Analysis

```{r trajectories, fig.height=6}
set.seed(123)
sample_ids <- sample(unique(longitudinal_data$id), 20)

longitudinal_data %>%
  filter(id %in% sample_ids) %>%
  left_join(survival_data %>% select(id, status), by = "id") %>%
  ggplot(aes(x = time, y = observed, group = id, color = factor(status))) +
  geom_line(alpha = 0.6) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(
    values = c("0" = "#3498db", "1" = "#e74c3c"),
    labels = c("Censored", "Event"), name = "Status"
  ) +
  labs(title = "Log Serum Bilirubin Trajectories (20 Patients)",
       x = "Time (years)", y = "Log Bilirubin (standardized)") +
  theme_minimal(base_size = 12)
```

## Mean Trajectories by Event Status

```{r mean-trajectories}
longitudinal_data %>%
  left_join(survival_data %>% select(id, status), by = "id") %>%
  group_by(status, time_bin = cut(time, breaks = seq(0, 15, by = 0.5))) %>%
  summarise(
    mean_val = mean(observed),
    se_val = sd(observed) / sqrt(n()),
    time_mid = mean(time),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = time_mid, y = mean_val, color = factor(status))) +
  geom_line(linewidth = 1) +
  geom_ribbon(
    aes(ymin = mean_val - 1.96 * se_val,
        ymax = mean_val + 1.96 * se_val,
        fill = factor(status)),
    alpha = 0.2, color = NA
  ) +
  scale_color_manual(
    values = c("0" = "#3498db", "1" = "#e74c3c"),
    labels = c("Censored", "Event"), name = "Status"
  ) +
  scale_fill_manual(
    values = c("0" = "#3498db", "1" = "#e74c3c"),
    labels = c("Censored", "Event"), name = "Status"
  ) +
  labs(title = "Mean Log Bilirubin by Event Status",
       x = "Time (years)", y = "Mean Log Bilirubin (standardized)") +
  theme_minimal(base_size = 12)
```

# Model Fitting

## JointODE

```{r fit-jointode}
fit_ode <- JointODE(
  longitudinal_formula = observed ~ biomarker + velocity + drug +
    (biomarker + velocity | id),
  survival_formula = Surv(time, status) ~ drug + age + sex,
  longitudinal_data = longitudinal_data,
  survival_data = survival_data,
  control = list(parallel = TRUE, verbose = 3)
)
summary(fit_ode)
cat(sprintf("C-index: %.3f\n", fit_ode$cindex))
```

## Extended Cox (Time-Varying Covariates)

A Cox model with log serum bilirubin as a time-varying covariate using the counting process `(start, stop]` format with last-observation-carried-forward (LOCF).

```{r fit-extended-cox}
# Build (start, stop] intervals with LOCF biomarker
td <- longitudinal_data %>%
  arrange(id, time) %>%
  left_join(survival_data %>% select(id, time_event = time, status), by = "id") %>%
  group_by(id) %>%
  mutate(
    tstart = time,
    tstop = lead(time, default = first(time_event)),
    event = ifelse(row_number() == n(), first(status), 0L)
  ) %>%
  ungroup() %>%
  filter(tstop > tstart)

cox_fit <- coxph(
  Surv(tstart, tstop, event) ~ observed + drug + age + sex,
  data = td, x = TRUE
)
summary(cox_fit)
```

## JMbayes2

```{r fit-jmbayes2}
lme_fit <- lme(
  observed ~ ns(time, 3) * drug,
  random = ~ time | id,
  data = longitudinal_data
)

fit_jm <- jm(cox_fit, lme_fit, time_var = "time",
              functional_forms = list(
                "observed" = ~ value(observed) + slope(observed)
              ))

summary(fit_jm)
```

## Model Comparison

```{r model-comparison}
ode_c <- coef(fit_ode)
ode_se <- sqrt(diag(vcov(fit_ode)))
jm_s <- summary(fit_jm)$Survival
cox_s <- summary(cox_fit)$coefficients

fmt <- function(est, se) sprintf("%.3f (%.3f)", est, se)

comp <- data.frame(
  Parameter = c("observed (value)", "observed (slope)", "drug", "age", "sex"),
  Extended.Cox = c(
    fmt(cox_s["observed", "coef"], cox_s["observed", "se(coef)"]),
    "--",
    fmt(cox_s["drug", "coef"], cox_s["drug", "se(coef)"]),
    fmt(cox_s["age", "coef"], cox_s["age", "se(coef)"]),
    fmt(cox_s["sex", "coef"], cox_s["sex", "se(coef)"])
  ),
  JointODE = c(
    fmt(ode_c["hazard:alpha_1"], ode_se["hazard:alpha_1"]),
    fmt(ode_c["hazard:alpha_2"], ode_se["hazard:alpha_2"]),
    fmt(ode_c["hazard:drug"], ode_se["hazard:drug"]),
    fmt(ode_c["hazard:age"], ode_se["hazard:age"]),
    fmt(ode_c["hazard:sex"], ode_se["hazard:sex"])
  ),
  JMbayes2 = c(
    fmt(jm_s["value(observed)", "Mean"], jm_s["value(observed)", "StDev"]),
    fmt(jm_s["slope(observed)", "Mean"], jm_s["slope(observed)", "StDev"]),
    fmt(jm_s["drug", "Mean"], jm_s["drug", "StDev"]),
    fmt(jm_s["age", "Mean"], jm_s["age", "StDev"]),
    fmt(jm_s["sex", "Mean"], jm_s["sex", "StDev"])
  )
)

kable(comp, align = c("l", "r", "r", "r"),
      caption = "Hazard coefficients: estimate (SE)")
```
# Session Info

```{r session-info}
sessionInfo()
```
