---
title: "Illustration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, echo=FALSE}
library(JointODE)
library(dplyr)
library(purrr)

library(nlme)
library(ggplot2)
library(gridExtra)
library(grid)
library(latex2exp)
```

## Mathematical Framework

```{r function, echo=FALSE}
.solve_simple_biomarker_ode <- function(
  times = seq(0, 100, by = 0.1),
  init = c(1, 0),
  configurations = list()
) {
  .calculate_acceleration <- function(t, biomarker, velocity, parms) {
    omega <- 2 * pi / parms$period
    -2 *
      as.numeric(parms$xi * omega) *
      velocity -
      as.numeric(omega^2) * biomarker +
      parms$k * as.numeric(omega^2) * parms$excitation(t)
  }
  .ode_deriv <- function(t, state, parms) {
    biomarker <- state[1]
    velocity <- state[2]
    acceleration <- .calculate_acceleration(
      t,
      biomarker,
      velocity,
      parms
    )
    list(c(velocity, acceleration))
  }
  default_config <- list(
    xi = -0.2,
    period = 20,
    k = 1.0,
    excitation = function(t) 0
  )
  ode_parms <- modifyList(default_config, configurations)
  ode_solution <- deSolve::ode(
    y = init,
    times = sort(c(0, times)),
    func = .ode_deriv,
    parms = ode_parms
  )
  acceleration <- .calculate_acceleration(
    ode_solution[, 1],
    ode_solution[, 2],
    ode_solution[, 3],
    ode_parms
  )
  idx <- match(times, ode_solution[, 1])
  data.frame(
    time = times,
    biomarker = ode_solution[idx, 2],
    velocity = ode_solution[idx, 3],
    acceleration = acceleration[idx]
  )
}
```


$$\ddot{m}(t) + 2\xi\omega\dot{m}(t) + \omega^2 m(t) = k\omega^2 u(t), \quad m(0) = m_0, \quad \dot{m}(0) = v_0$$

**Parameters:**

- $m(t)$: biomarker value
- $\xi$: damping ratio (stability control)
- $\omega = 2\pi/T$: natural frequency
- $k$: steady-state gain
- $u(t)$: external excitation

### Systematic Classification

Complete 2*3 matrix: $\omega^2$ (motion type) * $\omega\xi$ (energy behavior)

```{r dynamic-cases, fig.width=12, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
cases <- list(
  # omega^2 < 0 (all unstable)
  list(
    name = "omega^2 < 0, omega*xi < 0",
    xi = 0.5i, period = 20i, k = 1.0,
    excitation = function(t) 0, init = c(1, 0)
  ),
  list(
    name = "omega^2 < 0, omega*xi = 0",
    xi = 0, period = 20i, k = 1.0,
    excitation = function(t) 0, init = c(1, 0)
  ),
  list(
    name = "omega^2 < 0, omega*xi > 0",
    xi = -0.5i, period = 20i, k = 1.0,
    excitation = function(t) 0, init = c(1, 0)
  ),

  # omega^2 > 0
  list(
    name = "omega^2 > 0, omega*xi < 0",
    xi = -0.3, period = 20, k = 1.0,
    excitation = function(t) 0, init = c(1, 0)
  ),
  list(
    name = "omega^2 > 0, omega*xi = 0",
    xi = 0, period = 20, k = 1.0,
    excitation = function(t) 0, init = c(1, 0)
  ),
  list(
    name = "omega^2 > 0, omega*xi > 0",
    xi = 0.3, period = 20, k = 1.0,
    excitation = function(t) 0, init = c(1, 0)
  )
)

# Simulate each case
case_data <- map_df(cases, function(case) {
  # Shorter time for unstable cases
  max_time <- 100
  times <- seq(0, max_time, by = 0.1)

  result <- .solve_simple_biomarker_ode(
    times = times,
    init = case$init,
    configurations = list(
      xi = case$xi,
      period = case$period,
      k = case$k,
      excitation = case$excitation
    )
  )

  # Calculate ω and display values
  omega <- 2 * pi / case$period
  omega_squared <- omega^2 * case$k  # Effective omega^2
  omega_xi <- omega * case$xi

  data.frame(
    time = result$time,
    biomarker = result$biomarker,
    velocity = result$velocity,
    case_name = case$name,
    omega_squared = as.numeric(omega_squared),
    omega_xi = as.numeric(omega_xi),
    xi = as.numeric(case$xi)
  )
})

# Time series plot
p_cases_time <- ggplot(
  case_data,
  aes(x = time, y = biomarker)
) +
  geom_line(linewidth = 1, alpha = 0.9, aes(color = omega_squared)) +
  geom_hline(yintercept = 0, linetype = "dotted", alpha = 0.4) +
  facet_wrap(~case_name, scales = "free", ncol = 3,
             labeller = labeller(case_name = function(x) {
               gsub("omega\\^2", "\u03C9\u00B2", gsub("omega\\*xi", "\u03C9\u03BE", x))
             })) +
  scale_color_gradient2(
    name = expression(omega^2),
    low = "red",
    mid = "gray",
    high = "blue",
    midpoint = 0
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 9),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "Time",
    y = "Biomarker Value"
  )

# Display biomarker dynamics
print(p_cases_time)

# Phase portrait
p_cases_phase <- ggplot(
  case_data,
  aes(x = biomarker, y = velocity)
) +
  geom_path(linewidth = 0.8, alpha = 0.7, aes(color = time)) +
  geom_point(
    data = case_data %>%
      group_by(case_name) %>%
      slice_min(time),
    size = 3, shape = 21, fill = "green", color = "darkgreen"
  ) +
  facet_wrap(~case_name, scales = "free", ncol = 3,
             labeller = labeller(case_name = function(x) {
               gsub("omega\\^2", "\u03C9\u00B2", gsub("omega\\*xi", "\u03C9\u03BE", x))
             })) +
  scale_color_viridis_c(name = "Time", option = "magma") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 9),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Phase Portraits",
    x = "Biomarker",
    y = "Velocity (dm/dt)"
  )

# Display phase portraits
print(p_cases_phase)
```

### Autonomous Systems

**Constant excitation** ($u(t) = 0$):

```{r simulation, echo=FALSE, message=FALSE, warning=FALSE}
# Define damping ratios including key values
xi <- c(-0.1, 0, 0.1, 0.3, 0.5, 0.707, 0.9, 1.0, 1.2, 1.5, 2.0)

# Categorize damping ratios
xi_group <- cut(
  xi,
  breaks = c(-Inf, 0, 0.7, 1, Inf),
  labels = c(
    "Divergent (xi <= 0)",
    "Underdamped (0 < xi <= 0.7)",
    "Critical (0.7 < xi < 1)",
    "Overdamped (xi >= 1)"
  ),
  include.lowest = TRUE,
  right = TRUE
)

# Run simulations
simulations <- map(
  xi,
  ~ .solve_simple_biomarker_ode(configurations = list(xi = .x))
)

# Combine results
simulation_data <- map2_df(
  simulations,
  seq_along(simulations),
  function(data, idx) {
    data.frame(
      time = data$time,
      biomarker = data$biomarker,
      velocity = data$velocity,
      damping_ratio = xi[idx],
      group = xi_group[idx]
    )
  }
)
```

```{r dynamics-plots, fig.width=12, fig.height=6, echo=FALSE}
# Time series plot
p_time <- ggplot(
  simulation_data,
  aes(x = time, y = biomarker, color = as.factor(damping_ratio))
) +
  geom_line(linewidth = 1, alpha = 0.85) +
  geom_hline(yintercept = 0, linetype = "dotted", alpha = 0.4) +
  facet_wrap(~group, scales = "free_y", ncol = 2) +
  scale_color_viridis_d(name = expression(xi), option = "plasma") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    plot.title = element_text(face = "bold", size = 12),
    strip.text = element_text(face = "bold", size = 9),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank()
  ) +
  labs(x = "Time", y = "Biomarker")

# Phase portrait plot
selected_xi <- c(-0.1, 0.3, 0.707, 1.5)
phase_data <- simulation_data %>%
  filter(damping_ratio %in% selected_xi)

p_phase <- ggplot(phase_data, aes(x = biomarker, y = velocity)) +
  geom_path(aes(color = time), linewidth = 1, alpha = 0.8) +
  geom_point(
    data = phase_data %>%
      group_by(damping_ratio) %>%
      slice_min(time),
    size = 2, shape = 21, fill = "green", color = "darkgreen"
  ) +
  geom_point(
    data = phase_data %>%
      group_by(damping_ratio) %>%
      slice_max(time),
    size = 2, shape = 21, fill = "red", color = "darkred"
  ) +
  facet_wrap(~ damping_ratio, scales = "free", ncol = 2,
             labeller = labeller(damping_ratio = function(x) paste0("\u03BE = ", x))) +
  scale_color_viridis_c(name = "Time", option = "magma") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 12),
    strip.text = element_text(face = "bold", size = 9),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "Biomarker",
    y = "Rate of Change (dm/dt)",
    caption = "Initial conditions (green) and terminal states (red) indicated"
  )

# Combine plots side by side
grid.arrange(p_time, p_phase, ncol = 2, widths = c(1, 1))
```

- **Stable:** $\omega^2 \geq 0$ AND $\omega\xi > 0$
- **Marginally stable:** $\omega^2 \geq 0$ AND $\omega\xi = 0$
- **Unstable:** $\omega^2 < 0$ OR $\omega\xi < 0$

### Non-Autonomous Systems

**Step excitation:** $u(t) = 0.5 \cdot \mathbb{1}_{t \geq 50}$

```{r non-autonomous, fig.width=12, fig.height=6, echo=FALSE}
# Use same xi values as autonomous case
# Define step input function
step_input <- function(t) {
  ifelse(t < 50, 0, 1) * 0.5
}

# Simulate responses using JointODE function
na_data <- map2_df(xi, xi_group, function(xi_val, group) {
  result <- .solve_simple_biomarker_ode(
    configurations = list(xi = xi_val, excitation = step_input)
  )

  data.frame(
    time = result$time,
    biomarker = result$biomarker,
    velocity = result$velocity,
    damping_ratio = xi_val,
    group = group
  )
})

# Time series plot
p_time_na <- ggplot(
  na_data,
  aes(x = time, y = biomarker, color = as.factor(damping_ratio))
) +
  geom_line(linewidth = 1, alpha = 0.85) +
  geom_vline(
    xintercept = 50, linetype = "dashed",
    color = "red", alpha = 0.5
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", alpha = 0.4) +
  geom_hline(yintercept = 1, linetype = "dotted", alpha = 0.4) +
  facet_wrap(~group, scales = "free_y", ncol = 2) +
  scale_color_viridis_d(name = expression(xi), option = "plasma") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    plot.title = element_text(face = "bold", size = 12),
    strip.text = element_text(face = "bold", size = 9),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank()
  ) +
  labs(x = "Time", y = "Biomarker")

phase_data_na <- na_data %>%
  filter(damping_ratio %in% c(-0.1, 0.707, 1.0, 1.5))

p_phase_na <- ggplot(phase_data_na, aes(x = biomarker, y = velocity)) +
  geom_path(aes(color = time), linewidth = 1, alpha = 0.8) +
  geom_point(
    data = phase_data_na %>%
      group_by(damping_ratio) %>%
      slice_min(time),
    size = 2, shape = 21, fill = "green", color = "darkgreen"
  ) +
  geom_point(
    data = phase_data_na %>%
      group_by(damping_ratio) %>%
      slice_max(time),
    size = 2, shape = 21, fill = "red", color = "darkred"
  ) +
  facet_wrap(~ damping_ratio,
    scales = "free", ncol = 2,
    labeller = labeller(damping_ratio = function(x) paste0("\u03BE = ", x))
  ) +
  scale_color_viridis_c(name = "Time", option = "magma") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 12),
    strip.text = element_text(face = "bold", size = 9),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "Biomarker",
    y = "Rate of Change (dm/dt)",
    caption = "Initial conditions (green) and terminal states (red) indicated"
  )

# Combine plots side by side
grid.arrange(p_time_na, p_phase_na, ncol = 2, widths = c(1, 1))
```

### Real Data Examples

Four biological datasets illustrate distinct damping behaviors:

- **Canadian Lynx** (1821-1934): Annual trapping data showing predator-prey population cycles with pronounced oscillations
- **Theophylline**: Pharmacokinetic data from 12 subjects showing drug concentration after oral administration of the anti-asthmatic medication
- **Indomethacin**: Plasma concentration following intravenous injection of the anti-inflammatory drug in 6 subjects
- **Orange Tree Growth**: Trunk circumference measurements over time from a forestry growth study

```{r real-data, fig.width=12, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
# Prepare data
data(lynx)
lynx_df <- data.frame(time = 1821:1860, value = lynx[1:40])

data(Theoph)
theoph_df <- Theoph[Theoph$Subject == 1, ]
theoph_df <- data.frame(time = theoph_df$Time, value = theoph_df$conc)

data(Indometh)
indo_df <- Indometh[Indometh$Subject == 1, ]
indo_df <- data.frame(time = indo_df$time, value = indo_df$conc)

data(Orange)
orange_df <- Orange[Orange$Tree == 1, ]
orange_df <- data.frame(time = orange_df$age, value = orange_df$circumference)

# Create individual plots with specific axis labels
p_lynx <- ggplot(lynx_df, aes(x = time, y = value)) +
  geom_line(linewidth = 1, alpha = 0.9, color = "#C73E1D") +
  geom_point(size = 1.5, alpha = 0.7, color = "#C73E1D") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8)
  ) +
  labs(
    title = "Lynx",
    x = "Year",
    y = "Population (thousands)"
  )

p_theoph <- ggplot(theoph_df, aes(x = time, y = value)) +
  geom_line(linewidth = 1, alpha = 0.9, color = "#2E86AB") +
  geom_point(size = 1.5, alpha = 0.7, color = "#2E86AB") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8)
  ) +
  labs(
    title = "Theophylline", x = "hours", y = "mg/L"
  )

p_indo <- ggplot(indo_df, aes(x = time, y = value)) +
  geom_line(linewidth = 1, alpha = 0.9, color = "#A23B72") +
  geom_point(size = 1.5, alpha = 0.7, color = "#A23B72") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8)
  ) +
  labs(
    title = "Indomethacin", x = "hours", y = "mcg/ml"
  )

p_orange <- ggplot(orange_df, aes(x = time, y = value)) +
  geom_line(linewidth = 1, alpha = 0.9, color = "#F18F01") +
  geom_point(size = 1.5, alpha = 0.7, color = "#F18F01") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8)
  ) +
  labs(title = "Orange Tree", x = "days", y = "Circumference (mm)")

# Combine plots
grid.arrange(p_lynx, p_theoph, p_indo, p_orange, ncol = 2, nrow = 2)
```

This analysis requires $\omega^2 \geq 0$ for real natural frequencies, focusing on stable or oscillatory systems that better represent biological reality. Systems with $\omega^2 < 0$ exhibit purely exponential divergence without oscillations and lack practical relevance for modeling physiological processes.

### Theoretical Analysis

| Damping | Condition | Solution Form | Behavior |
|---------|-----------|---------------|----------|
| Underdamped | $0 < \xi < 1$, $\omega^2 > 0$ | $e^{-\xi\omega t}[A\cos(\omega_d t) + B\sin(\omega_d t)]$ | Decaying oscillations |
| Critically damped | $\xi = 1$, $\omega^2 > 0$ | $(A + Bt)e^{-\omega t}$ | Fastest decay without oscillation |
| Overdamped | $\xi > 1$, $\omega^2 > 0$ | $Ae^{\lambda_1 t} + Be^{\lambda_2 t}$ | Slow exponential decay |
| Undamped | $\xi = 0$, $\omega^2 > 0$ | $A\cos(\omega t) + B\sin(\omega t)$ | Sustained oscillations |
| Unstable | $\omega^2 < 0$ or $\omega\xi < 0$ | Exponential growth | System diverges |

where:

- $\omega_d = \omega\sqrt{1 - \xi^2}$: damped natural frequency
- $\lambda_{1,2} = -\xi\omega \pm \omega\sqrt{\xi^2 - 1}$: overdamped eigenvalues
- $A, B$: constants determined by initial conditions $m_0, v_0$

## Joint Model with Survival Outcomes

The hazard function incorporates both the biomarker level and its rate of change:

$$\lambda(t) = \lambda_0(t) \exp(\alpha_1 m(t) + \alpha_2 \dot{m}^{\gamma}(t))$$

where:

- $\lambda_0(t)$: baseline hazard function
- $\alpha_1, \alpha_2$: regression coefficients for biomarker level and velocity
- $\gamma=0,1,2$: power transformation for velocity (e.g., $\gamma=2$ squared to capture magnitude)

This formulation captures two distinct aspects of risk:

- $m(t)$: The biomarker level itself (e.g., elevated glucose indicating immediate risk)
- $\dot{m}(t)$: The biomarker volatility (e.g., rapid fluctuations indicating instability)

By incorporating $\dot{m}^{\gamma}(t)$, we effectively model both the biomarker's central tendency and its temporal variability—analogous to simultaneously capturing mean and variance effects on survival risk. This dual characterization provides richer prognostic information than static measurements alone.

**Remark:** For oscillatory biomarker dynamics ($\xi < 1$), using absolute values prevents positive and negative velocity contributions from canceling in cumulative hazard calculations, ensuring that all variability contributes to risk assessment.
